---
title: "Lab5"
author: "Laura Lázaro Soraluce"
date: "2025-03-26"
output: html_document
---

### Implementation in R: The case of known derivative
```{r}
f<-function(x) x^2-5
curve(f,0,10)
abline(h=0, col=2)

df<-function(x) 2*x
x0<-2
x1<-x0-f(x0)/df(x0); x1 #First iteration
x2<-x1-f(x1)/df(x1); x2 #Second iteration
x3<-x2-f(x2)/df(x2); x3 #Third iteration
x4<-x3-f(x3)/df(x3); x4 #Fourth iteration
x5<-x4-f(x4)/df(x4); x5 #Fifth iteration

# Function that computes the desired n values
newton_r<-function(x0, f, df, n) {
  x<-numeric(n)
  x<-x0
  for(i in 2:(n+1)) {
    x[i]<-x[i-1]-f(x[i-1])/df(x[i-1])
  }
  return (x)
}

#f(x)=x^3-2x-5
f2<-function(x) x^3-2*x-5
curve(f2,-5,5)
abline(h=0,col=2)
```

##### We can see in the curve that the root is close to 2, so we use that as the initial value

```{r}
# We compute a few iterations
df2<-function(x) 3*x^2-2
newton_r(2, f2, df2, 5)
```

##### Once again, we can see that it reaches the root by the third iteration

```{r}
# We try with a poor initial value
newton_r(0, f2, df2, 5)

# Let's add a few more iterations
newton_r(0, f2, df2, 10)
```

##### Even if we add a few more iterations, it diverges, as the values move up and down, but nowhere around the real value

```{r}
# Let's try a different initial value
newton_r(-3, f2, df2, 7)
newton_r(4, f2, df2, 7)
```

##### In both of these cases, it does converge by the 7th and 6th iteration respectivelly. This suggests that it doesn't converge when then initial values are in the flat region of the function, since we're using the derivative and it's very close to zero, causing big flunctuations.


##### I had already implemented a function, so I'll adapt it to what the worksheet says
```{r}
NR<-function(x0, f, df, nmax=50, tol=1e-3) {
  x<-x0
  i<-0
  
  while(i<nmax){
    
    if (df(x)==0){
      stop("Derivative is 0")
    }
    xnew<-x-f(x)/df(x)
    
    final.diff = abs(xnew - x)
    if(final.diff<tol){
      return (list(root = xnew, final.diff, niter=i+1))
    }
    
    x<-xnew
    i<-i+1
  }
  
  warning("Max number of iterations reached without converging")
  return (list(root = xnew, final.diff, niter=nmax))
}

NR(2,f,df)
NR(2,f2,df2)
```

##### As we can see, the new function works perfectly

### Implementarion using numerical differentiation
```{r}
library(numDeriv)

f<-function(x) x^2-5
res<-genD(func=f, x=2)
res$D[1]

x0<-2
x1<-x0-f(x0)/genD(func=f,x=x0)$D[1]; x1
```

```{r}

NR2<-function(x0, f, df, nmax=50, tol=1e-5) {
  x<-x0
  i<-0
  
  while(i<nmax){
    if(missing(df)){
      deriv<-genD(func=f, x=x)$D[1]
      xnew<-x-f(x)/deriv
    }
    else {
      if (df(x)==0){
        stop("Derivative is 0")
      }
      xnew<-x-f(x)/df(x)
    }
    final.diff = abs(xnew - x)
    if(final.diff<tol){
      return (list(root = xnew, final.diff, niter=i+1))
    }
    
    x<-xnew
    i<-i+1
  }
  
  warning("Max number of iterations reached without converging")
  return (list(root = xnew, final.diff, niter=nmax))
}

NR2(2, f)
NR2(2, f2)
f3<-function(x) (exp(2*x)-x-6)
curve(f3,5,10)
abline(h=0,col=2)
NR2(8,f3)

uniroot(f, c(0,4), tol=1e-5)$root
uniroot(f2, c(0,4), tol=1e-5)$root
uniroot(f3, c(0,4), tol=1e-5)$root

```

##### We obtain almost the same results, just a tiny difference in the last decimals.

```{r}
find_root<-function(f, a0, b0, nmax=100, tol=1e-5){
  i<-1
  a<-a0
  b<-b0
  while (i<nmax){
    r<-a+((b-a)/2)
    if (f(r)==0 || ((b-a)/2)<tol ){
      return(list(root=r, niter=i))
    }
    i<-i+1
    if((f(a)*f(r))>0){
      a=r
    }
    else {
      b=r
    }
  }
  warning("Max number of iterations reached without converging")
  return (list(root = r, niter=nmax))
}

f4<-function(x) x^3+4*x^2-10

# Find_root
find_root(f4, 1, 2)
# Uniroot
uniroot(f4, c(1,2), tol=1e-5)$root
# NR2
NR2(1.5, f4)
```

##### The results we get using these 3 functions are basically the same, they just differ in the last decimal. However, the number of iterations until the root value is obtained, is different for the methods find_root and NR2, being a lot smaller for this last one.
