---
title: "Lab4"
author: "Laura"
date: "2025-03-20"
output: html_document
---

```{r}
amounts<-scan(text="26.55 19.87 22.62 27.53 3.76 1.79 13.43 23.76 58.58 64.74
9.87 50.59 25.79 80.11 59.19 15.08 7.19 41.25 14.09 24.74
4.59 70.40 82.11 18.90 5.46 252.06 50.40 51.32 21.96 28.29
215.15 93.02 45.92 70.48 64.81 1.57 48.24 22.84 204.80 134.66
120.93 8.38 19.16 18.99 90.66 24.63 27.24 90.91 82.11 37.85")

amounts

```
### Method 1
```{r}
##### To see that it works properly
a<-1; b<-1
print(dgamma(x=amounts, shape=a, scale=b))

logl<-function(theta)
{
  a<-theta[1]
  b<-theta[2]
  l<-sum(log(dgamma(x=amounts,shape=a,scale=b)))
  return(-l)
}

logl(c(1,1))
logl(c(2.5,1))
logl(c(0,2))
logl(c(1,-1))
```
##### As we can see, we get a warning in the last case, since we're using a negative value for b, and it expects a positive one. In the 3rd case, we get Inf since we're using a 0 and a has to be positive.

```{r}
res<-optim(par=c(1,1),fn=logl)
res$par

##### We inspect the object res
str(res)

##### We now try with different initial values
res1<-optim(par=c(2,2),fn=logl)
res1$par
res2 <- optim(par=c(0.5,0.5), fn=logl)
res2$par
```
##### The solution isn't really sensitive to the initial values because even if we change them, we obtain almost the same optimum (really small fluctuations).

```{r}
##### Optimization using solnp
library(Rsolnp)

b_m<-var(amounts)/mean(amounts)
a_m<-mean(amounts)/b_m

a_m
b_m
```
##### In this case, we get a_m = 0.916 and b_m=56.72, as opposed to the results we got before: a=1.102, b=47.1. They are not the exact same, but fairly similar. I would say the Maximum Likelihood method is more precise, as it doesn't simply match moments. 
```{r}
res_solnp<-solnp(pars=c(a_m,b_m), fun=logl, LB=c(0,0))
```
##### We have obtained very similar values to the ones we got with Maximum Likelihood again. This optimization works greatly, as it has converged to the optimum in only two iterations. 

```{r}
#####  Optimization using maxLik
library(maxLik)

logl2<-function(theta) -logl(theta)
A<-matrix(c(1,0,0,1),2)
B<-c(0,0)
maxLik(logl2,start=c(a,b),constraints=list(ineqA=A,ineqB=B))
```

### Method 2
```{r}
f <- function(a) log(a) - digamma(a) - log(mean(amounts)) + mean(log(amounts))
res <- uniroot(f, c(0.1, 100))
res

##### We compute b from a
b_uniroot <- mean(amounts) / res$root
b_uniroot
```
##### We get almost the same result using Maximum Likelihood and the normal equations, so both are reliable. And this last result is, again, a bit different than the one obtained with the moment estimators. 

### Proposed exercise
```{r}
amounts2<- scan(text="0.97 1.01 1.06 0.99 1.13 1.08 0.98 0.93 1.03 0.97
0.83 1.07 1.06 0.49 0.84 1.07 1.01 0.81 1.01 0.90
0.93 0.98 1.01 0.77 0.90 0.93 0.88 1.08 1.03 1.09
0.88 0.91 0.89 0.98 0.91 0.90 0.85 0.91 0.90 0.81
1.01 0.91 1.00 1.04 0.81 0.86 0.94 0.69 1.01 0.86")

##### logl function
weibull_logl <- function(params) {
  a <- params[1]
  sigma <- params[2]

  n <- length(amounts2)
  log_likelihood <- n * log(a) - n * a * log(sigma) + (a - 1) * sum(log(amounts2)) - sum((amounts2 / sigma)^a)
  
  return(-log_likelihood)
}

##### optim
weibull_mle <- optim(par = c(a, b), fn = weibull_logl)

a_mle <- weibull_mle$par[1]
sigma_mle <- weibull_mle$par[2]

a_mle
sigma_mle

##### solnp
weibull_solnp <- solnp(pars = c(1,1), fun = weibull_logl, LB = c(0.0, 0.0))

a_solnp <- weibull_solnp$pars[1]
sigma_solnp <- weibull_solnp$pars[2]

a_solnp
sigma_solnp

##### uniroot
weibull_eq <- function(a) {
  sum_x_a <- sum(amounts2^a)
  sum_x_a_logx <- sum(amounts2^a * log(amounts2))
  sum_logx <- sum(log(amounts2))
  n <- length(amounts2)
  return(sum_x_a_logx/sum_x_a - 1/a - sum_logx/n)
}

a_root <- uniroot(weibull_eq, c(0.1, 20))$root
sigma_root <- (sum(amounts2^a_root) / length(amounts2))^(1/a_root)

a_root
sigma_root

```
##### We get the same results for all the methods, which proves that they all find the global maximum and are solving the normal equations correctly. Solnp converges in 2 iterations, so it's effective, once again. 
